## Advanced R {-}

[Advanced R](https://adv-r.hadley.nz/) 2nd ed. Hadley Wickham

### Names and Values

#### Binding {-}
In R, names are assigned a value (e.g. a vector doesn't have a name, rather, a name(s) points to a vector). The actual address of an object is exposed by `lobstr::obj_addr()`.
```{r include=FALSE}
library(lobstr)
```
```{r echo=TRUE, collapse=TRUE}
x <- 1:10
obj_addr(x)

# notice x and y are pointers to the same object
y <- x
obj_addr(y)

# this applies to function arguments in their environment as well
fn <- function(z){
  z
}
# fn() returns the argument z at the same address
obj_addr(fn(x))
```

#### Syntactic Names {-}

Names in R can be letters, numbers, `.` or `_`; but cannot start with a number or `_` or contain `?Reserved` words.

```{r echo=TRUE, collapse=TRUE, error=TRUE}
# assigning a vector a name startign with `_` throws an error
_x <- 1:10
```


#### Copy-on-Modify {-#copyonmod}

R objects are generally immutable. A new copy is made when you modify an object. There are two [Modify-in-Place](#modinplace) exceptions.

```{r, echo=TRUE, collapse=TRUE}
# create vector x
x <- 1:10
obj_addr(x)

# modify first value in x
x[[1]] <- 0

# note x now points to a new object
obj_addr(x)
```


##### Trace Copying of Objects {-}

`base::tracemem()` will mark an object and print a message whenever it is copied. This is
a major cause of hard-to-predict memory usage.

```{r echo=TRUE, eval=FALSE, collapse=TRUE}
# tell R to trace copies of object reference `x`
tracemem(x = x)

# modify the first element in vector `x`
# R prints a tracemem message to show the object is copied to a new address
x[[1]] <- 2
#> tracemem[0x000002379ca5a190 -> 0x000002379cd8dbb0]:

# tell R to stop tracing this object
base::untracemem(x = x)
```

##### List Objects {-#lobjects}

Just like variables, each element of a list also points to a value. [Copy-on-modify](#copyonmod) and [modify-in-place](#modinplace) applies here as well. R creates a **shallow** copy of the list. Meaning 
the list object and its bindings are copied; however, the underlying values are not.

In a deep copy, like prior to R 3.1.0, the underlying values are also copied.

```{r echo=TRUE, collapse=TRUE, }
# create a list named `l`
l <- list(1:10, TRUE, c('Apple','Broccoli','Chowder'))

# show the address of the third list element
obj_addr(l[[3]])

# modify the third list element
l[[3]] <- c('Appricot','Basmati Rice','Cheese')

# the third list element's address is changed
obj_addr(l[[3]])

```

Use `lobstr::ref()` to see common values between lists:

```{r echo=TRUE, collapse=TRUE}
# create a list object
listA <- list(1,2,3)

# create a new pointer to the same list object
listB <- listA

# modify the third list element of listB, creating a shallow copy of the list object 
listB[[3]] <- 4

# ref() lists the address of each list element. Notice two values are still shared 
# between `listA` and `listB`.
ref(listA, listB)

```

##### Data Frames {-}

A data frame is simply a list where each element is a vector of the same length.

```{r each=TRUE, collapse=TRUE}
# create a data frame with three columns
df <- data.frame(col1 = TRUE, 
                 col2 = 1:10, 
                 col3 = rep(x = c('On','Off'), 
                            length.out = 10))

# show the address of the data frame and each element (i.e. column)
ref(df)


```

When a **column** is modified only one element is copied:

```{r echo=TRUE, collapse=TRUE}
# modify a column and note the addresses of the unchanged columns remain the same
# because they point to the same objects as before
df$col1 <- FALSE; ref(df)
```

This has important consequences for memory when you update **rows** of data where each 
element gets copied:

```{r echo=TRUE, collapse=TRUE}
# modify the first row of the data frame
df[1,] <- c(TRUE, 0, 'Off')

# *all* elements are copied to new addresses
ref(df)

```

##### Character Vectors {-}

R uses a *global string pool* to store unique character objects so they are not duplicated 
unnecessarily. Use the `character` argument to show their address in the global string pool 
with `lobstr::ref()`:

```{r}
# create a character vector
chr <- c('Quarter',
         'Dime',
         'Nickle',
         'Penny',
         'Penny')

# show addresses in the global string pool. Notice the shared value for 'Penny'
ref(chr, 
    character = TRUE)

```

#### Object Size {-}

`lobstr::obj_size()` shows the amount of memory an object takes, while `obj_sizes()`
breaks down multiple objects into their individual contribution to *total* memory.

Due to binding names, global string pools, and ALTREP (alternate representation where some vectors
are stored in a compact manor; e.g. 1:1000 only stores the first and last numbers, not 1,000 numbers)
the size of objects may surprise you.

```{r echo=TRUE, collapse=TRUE}
# create a length one chracter vector and a length 100 character vector
x <- 'Charlie'
y <- rep(x = 'Charlie', times = 100)

# y is only ~8x the size of x, not 100x
obj_size(x); obj_size(y)


z <- 1:10

# create a list with one element `z`
list1 <- list(z)
# create a list with three elements, all `z`
list2 <- list(z, z, z)

# list2 only contributes an extra 80 Bytes to the total memory between the two
obj_sizes(list1, list2)

```


#### Modify-in-Place {-#modinplace}

There are two places where R will optimize memory by modfiying an object "in place"
(i.e. does not make a copy).

1. objects with a single binding

```
> hats <- c('red','brown','blue')
> obj_addr(hats)
[1] "0x1d51fe614d8"
> 
> hats[[1]] <- 'orange'
> obj_addr(hats)
[1] "0x1d51fe614d8"

# note this is output from the R Console. RStudio runs everything within an Environment
# which breaks modify-in-place behavior
```
This is one reason why looping in R is inefficient, the user is unwittingly copying 
objects many times over:

```{r echo=TRUE, collapse=TRUE}
# create a dataframe
df <- data.frame(col1 = c(1:3))

# trace copies of this dataframe
tracemem(df)

# use a for loop to increment the dataframe values
# R copies the object 12 times!!!
for (i in 1:3){
  df[[1]][i] <- df[[1]][i] + 1
}

# turn of object memory trace
untracemem(df)
```

2. Environments

Environments are always modified in place and all objects within the environment
keep the same reference.

#### Unbinding and Garbage Collector (GC) {-}

The (GC) garbage collector deletes objects that are no longer used and requests
more memory from the operating system as needed to create objects.

For example:
```{r echo=TRUE, collapse=TRUE}
# create three objects all bound to `x`
x <- 1:10 
x <- c('one','two','three')
x <- TRUE

# removing `x` causes GC to delete the three objects b/c they have no other bindings
remove(x)
```
You can call GC yourself with `base::gc()`, but the user should not ever have the need.

#### Session Info {-}

```{r echo=FALSE}
print(x = sessionInfo(), 
      local = F)
```
