```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, collapse=TRUE)
rm(list=ls())
```
## Misc Temp R

### combining execution

`{` curly braces can be used to execute multiple expressions; and only returns 
the last expression. It does not create a new environment outside of functions.

Separate statements with `;` or newline:

```{r}
{
  x <- 1
  y <- 2
  x+y
}

{
  set.seed(7); rnorm(1)
}
```

### Infix functions{#infixfuncs}

Operator | Type | Description | Note | Rank
--- | --- | --- | --- | ---
`::` | Extract | Retrieve Function | Extract function from package namespace; e.g. package::function() | 1
`:::` | Extract | Retrieve Function | Extract *hidden* function from namespace | 1
`$` | Extract | List Subset | Extract list element by __name__; see name() | 2
`@` | Extract | Slot selecton | Extract attributes by memory slot; see slotnames() | 2
`[` | Extract | Subscripting | Extract potentially **multiple** elements by index/dim vector; lists return lists | 3
`[[` | Extract | Subscripting | Extract **single** elements by index or name; lists return the object | 3
`^` | Arithmetic | Exponential | 2^3 = `r 2^3` | 4
`:` | Sequence | Sequence | 1:3 = `r 1:3`; 1:-1 = `r 1:-1` | 5
`%/%` | Arithmetic | Integer Divide | 5 %/% 2 = `r 5%/%2`; divide, then drop fractions | 6
`%%` | Arithmetic | Modulo | 7 %% 5 = `r 7%%5`; remainder after divison | 6
`%*%` | Arithmetic | Matrix Multiplication | Multiplies two conformable matricies | 6
`%o%` | Arithmetic | Outer Product | gives the outer product of arrays | 6
`%x%` | Arithmetic | Kronecker product | Kronecker product of two arrays | 6
`*` | Arithmetic | Multiplication | Also matrix dot product | 7
`/` | Arithmetic | Division | | 7
`+` | Arithmetic | Addition | | 8
`-` | Arithmetic | Subtraction | | 8
`!` | Comparison | Not | | 8
`%in%` | Comparison | Match Values | 1 %in% 1:3 = `r 1 %in% 1:3` | 9
`!=` | Comparison | Not equal to | 1 != 2 = `r 1 != 2` | 9
`<` | Comparison | Less than | 1 < 2 = `r 1 < 2` | 9
`>` | Comparison | Greater than | 1 > 3 = `r 1 > 3` | 9
`==` | Comparison | Equal to | 2 == 2 = `r 2==2` | 9
`<=` | Comparison | Less than or Equal to | 1 <= 1 = `r 1 <= 1` | 9
`>=` | Comparison | Greater than or Equal to | 2 >= 2 = `r 2 >= 2` | 9
`xor` | Logical | Exclusive or | element-wise if either, but not both, are TRUE: xor(1:2==1, 1:2<3) = `r xor(1:2==1, 1:2<3)` | 10
`&` | Logical | And (element) | element-wise logical AND
`&&` | Logical | And (control) | checks only first elements are logical AND; returns scalar | 10
`|` | Logical | Or (element) | element-wise logical OR | 10
`||` | Logical | Or (control) | checks only first elements are logical OR; returns scalar | 10
`~` | Assignment | Equal | Used for assignment in formulas | 11
`<<-` | Assignment | Permanent Assignment | Mainly used in closures to maintain state | 12
`<-` | Assignment | Left Assignment | | 13
`->` | Assignment | Right Assignment | | 13
`=` | Assignment | Argument Assignment | | 13

